Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes

### 2023-08-24 Journal-manuscript/Journal-manuscript.tex
Submitted this version to journal 1

% Repository:  https://github.com/chiehrosswang/TRB_LaTeX_tex
%
% Transportation Research Board conference paper template
% version 4.0 Lite (updates made to be compatible in Overleaf and ShareLaTeX)
%
%
% When numbered option is activated, lines are numbered.
\documentclass[]{trbunofficial}
\usepackage{graphicx}
\usepackage{booktabs}

\newread\somefile
\usepackage{xparse}
\usepackage{natbib}
\bibliographystyle{unsrtnat}
\setcitestyle{round}
% \usepackage[colorlinks=true,linkcolor=blue,citecolor=blue]{hyperref}
% For TRB version hide links
\usepackage[hidelinks]{hyperref}

% Put here what will go to headers as author
\AuthorHeaders{}
\title{Competition for Access to Opportunities between Modes: Multimodal
Spatial Availability}

% TODO: add macros for easier formatting of \author.
\author{%
    \textbf{Anastasia Soukhov}\\\textit{Corresponding Author}\\
  PhD Candidate\\
  Department of Earth, Environment and Society, McMaster University,
Canada\\
  \href{mailto:soukhoa@mcmaster.ca}{\nolinkurl{soukhoa@mcmaster.ca}}\\
  \hfill\break
    \textbf{Javier Tarriño-Ortiz}\\
  PhD Graduate\\
  Universidad Politécnica de Madrid, Spain\\
  \href{mailto:javier.tarrino.ortiz@upm.es}{\nolinkurl{javier.tarrino.ortiz@upm.es}}\\
  \hfill\break
    \textbf{Julio A. Soria-Lara}\\
  Associate Professor\\
  Urban and Regional Planning, Universidad Politécnica de Madrid,
Spain\\
  \href{mailto:julio.soria-lara@upm.es}{\nolinkurl{julio.soria-lara@upm.es}}\\
  \hfill\break
    \textbf{Antonio Páez}\\
  Associate Professor\\
  Department of Earth, Environment and Society, McMaster University,
Canada\\
  \href{mailto:paezha@mcmaster.ca}{\nolinkurl{paezha@mcmaster.ca}}\\
  \hfill\break
  }

% If necessary modify the number of words per table or figure default is set to
% 250 words per table (default defined in cls)

% If words are counted manually, put that number here. This does not include
% figures and tables. This can also be used to avoid problems with texcount
% program i.e. if one does not have it installed.
\TotalWords{7195}

@@

@@

\begin{document}
\maketitle

\section{Abstract}
An increasing number of studies within the domain of transportation
planning are concerned with the inequities in accessibility to
opportunities. A dimension of these inequities arises from differences
in access by mode type (e.g., commuting using a car as opposed to
transit). However, methods implemented in current accessibility
literature are lacking within the context of multimodal analysis. This
paper presents an extension of spatial availability, a
singly-constrained competitive accessibility measure, for the context of
multimodal accessibility analysis. We first illustrate the features of
spatial availability that lend itself to multimodal analysis. We then
demonstrate its use on the case study of Low Emission Zones in Madrid
(Spain) and highlight how this policy intervention changes the
accessibility of populations using different modes. In summary, spatial
availability can be used to create and interpret multimodal policy
intervention scenarios unlike previous methods: this creation and
interpretation can help regions envision a more sustainable and
equitable access-to-opportunity landscape.
\hfill\break%
\hfill\break%
\noindent\textit{Keywords}:  Multimodal, Accessibility, Equity, Policy
scenarios, Low emission zones,  
\newpage

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

An increasing number of studies within the domain of transport are
concerned with the inequities in accessibility to opportunities. A
dimension of these inequities arise from differences in access by mode
type (e.g., the number of work opportunities that can be reached using a
car as opposed to transit in a city). However, methods assessing
multimodal accessibility in the literature fall short as aspects of
competition for opportunities and the explicit methodological
acknowledgement of opportunities being \emph{finite} are lacking. In
this vein, this paper presents an extension of \emph{spatial
availability}, a singly-constrained competitive accessibility measure,
for the context of multimodal accessibility analysis. We first
illustrate the features of spatial availability that lends itself to
multimodal analysis. We then demonstrate its use on the case study of
Low Emission Zones in Madrid (Spain) and highlight how this policy
intervention changes the accessibility of populations using different
modes. In summary, spatial availability can be used to create and
interpret multimodal policy intervention scenarios unlike previous
methods: this creation and interpretation can help regions envision a
more sustainable and equitable access-to-opportunity landscape by better
identifying differences in accessibility afforded by different modes.

@@

Implementing urban policies that re-shape cities through accessibility
gains (i.e., the \emph{potential to interact} with opportunities as a
result of land-use mix and transport systems as originally defined by
\citet{hansenHowAccessibilityShapes1959}) have been widely applied
within the transportation literature and is increasingly discussed by
planners
\citep{levinsonTransportAccessManual2020, gowerPlanningInnovationCity2022, siddiqToolsTradeAssessing2021, yanAccessibilityBasedPlanningAddressing2021}.
An important challenge in the identification of interventions that
equitably transform cities is the effective evaluation of
\emph{trade-offs}: cities are complex and dynamic ecologies, and
advantaging one component of the city can disadvantage another area,
population, or sub-component In this way, policy evaluation should take
a \emph{systems} approach
\citep{fikselSustainabilityResilienceSystems2006}. One way of
considering systems is from the perspective of the \emph{finite}. As an
illustration, consider the amount of transport space within a city: the
amount is typically finite so re-allocating road space away from one
mode directly impacts the performance of the others (see the literature
on road space reallocation e.g.,
\citet{valencaMainChallengesOpportunities2021}). Evaluating policy
impacts in the context of \emph{finity} provides a way to contextualize
the balance of trade-offs that the citizens of a city should tolerate.

From the perspective of urban transport systems, location-based
accessibility measures have been used in the context of policy
evaluation. For instance, \citet{leeMeasuringImpactsNew2018} assesses
the transit accessibility gains to healthcare and employment
opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA
after the transit system's re-design and introduction of a rapid bus
system. However, a limitation of this study, like others that implement
accessibility measures, is they do not calculate results under a
\emph{constrained} framework i.e., one of \emph{finity}. The citizens of
Columbus should experience qualitative accessibility gains - but is it
at the expense of access to opportunities to populations using other
modes?

The work of \citet{leeMeasuringImpactsNew2018} uses
\emph{non-competitive} accessibility measures. There is a branch of
location-based accessibility measures that do incorporate the effect of
competition for opportunities by the population in the region. However,
we argue that these existing methods fall short in acknowledging the
\emph{finity} of opportunities. For instance,
\citet{maoMeasuringSpatialAccessibility2013} applies a competitive
measure, two-step floating catchment approach (2SFCA), for the case of
access to healthcare services in Florida for both a multimodal network
and a single modal network. While the differences in modal access are
discussed, the question of how the advantage in access afforded by one
mode over another impacts access for different mode users is unanswered.

This question of how much one mode-using population can access at the
expense of another mode-using population is a pertinent equity question
in the evaluation of policy scenarios that are multimodal. For instance,
consider the impact of a low emission zone (LEZ). LEZ is a policy of
spatial and modal discrimination: the circulation of vehicles that are
excessively polluting are restricted in specific regions. In the
recognition that opportunities are finite, the implementation of a LEZ
explicitly reduces the access that the population using polluting
vehicles has to opportunities. This restriction allows the population
using other more sustainable modes to potentially have a higher level of
access than before the LEZ implementation. This evaluation is especially
urgent as LEZ are currently in effect in cities globally; their
reception has been mixed \citep{tarrinoortizPublicAcceptabilityLow2021}
and may be having negative impacts on disadvantaged populations who have
become mobility restricted
\citep{devrijNooneVisitsMe2022, verbeekJustManagementUrban2022}.
Measures that evaluate the accessibility of modes given both
\emph{constrained} and \emph{competitive} considerations are lacking in
the literature, but are needed, to evaluate such policy interventions
impact on accessibility.

In \citet{soukhovIntroducingSpatialAvailability2023}, we introduce
spatial availability, a type of location-based accessibility measure
that is both \emph{constrained} and \emph{competitive}. In this paper,
we extend the spatial availability measure into a multimodal framework
and explore its use in answering the question outlined: \emph{``given
opportunities are finite, how many are available to a given location
depending on the mode used?''.} The answer to this question quantifies
how many opportunities can be accessed, considering competition, for
different modes. To foreground this exploration, in Section 2, we
discuss short falls of a few existing location-based measures in
comparison to spatial availability through a synthetic example. In
Section 3, the spatial availability of an empirical example of the LEZ
in the city of Madrid, Spain is calculated. We demonstrate how the
restriction of car circulation could have impacted the spatial
availability of opportunities for each sub-population using transit,
cycling and walking modes. In Section 4, we provide concluding remarks
on the strengths of the use of spatial availability as a multimodal
accessibility measure and potential future uses in policy planning
scenarios.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are a product of the relationship between land-use
and transport systems. Arguably the most commonly used location-based
measured are cumulative opportunity measures and weighted cumulative
opportunity measures \citep{levinsonTransportAccessManual2020}. These
measures weight the opportunities that can be potentially interacted
with from origin \(i\) to destination \(j\) based on some sort of travel
cost function (e.g., travel time, fare, travel distance) otherwise known
as a travel impedance function \(f^{m}(c^m_{ij})\). Many weighted
cumulative opportunities (often refered to as the gravity-based measure)
originate from the measure proposed by
\citet{hansenHowAccessibilityShapes1959}, which can take the following
multimodal form: \(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a
set of modes which have mode-specific travel costs \((c_{ij}^m)\) and
travel impedance functions \(f^m(\cdot)\).

The Hansen-type measure does not consider competition between modes nor
is it constrained. As an example, the work of
\citet{tahmasbiMultimodalAccessibilitybasedEquity2019} uses the
Hansen-type measure to measure the potential interaction with retail
locations using walking, public transit, and car modes \(m\). \(S_i^m\)
is the sum of retail locations \(j\) that can potentially be interacted
with under the travel impedance as calculated for each \(i\) and \(m\).
In other words, each \(i\) has three \(S_i\) values, one per \(m\). In
this work, they demonstrate that the car mode has the highest
\(S_i^{m=car}\) values in the majority of \(i\), i.e., populations using
a car can potentially interact with the most retail opportunities than
populations using other modes. However, the higher \(S_i^{m=car}\)
values are not a result of lower \(S_i^{m}\) values for other modes: it
is not assumed that car-using populations potentially accessing more
opportunities take away potential opportunities for other populations
within the measure (no consideration for competition). This measure is
also not constrained: there is no global maximum for \(S_i\) or
\(S_i^m\) values, they are presented as a population normalized
accessibility index. This makes the interpretation of the `potentially
interacted opportunities' relative to the region, making comparisons of
the results across different regions challenging.

However, opportunities in a region can be considered finite. There are
only so many school-seats, hospital capacity, jobs, etc., in a region
and if one person interacts with an opportunity at a given time, it is
taken. As such, if one person is advantaged and has the ability to reach
more opportunities through a lower travel-cost mode, than they have more
opportunities to potentially interact with more opportunities than other
people. From the other perspective, their are fewer opportunities left
to be potentially interacted with for populations using higher
travel-cost modes. In this way, populations using modes with a higher
travel impedance are at a higher access disadvantage than populations
using lower travel impedance modes. This recognition is the motivation
behind integrating \emph{competition} for opportunities within
multimodal accessibility measures. Arguably one of the most popular
competitive location-based accessibility measures is the two-step
floating catchment area (2SFCA) approach popularized by
\citet{luoMeasuresSpatialAccessibility2003} who simplified the approach
proposed by \citet{shenLocationCharacteristicsInnercity1998} (with
similar considerations for competition in
\citet{weibullAxiomaticApproachMeasurement1976} and
\citet{josephMeasuringPotentialPhysical1982}).

@@

To illustrate, \citet{taoInvestigatingImpactsPublic2020a} calculates
\(a_i^m\) to jobs for different income-group populations in Shenzhen
(China) using \(m = \text{public transit}\) and \(m={car}\). They
demonstrate that \(i\) with low-income populations have lower \(a_i^m\)
than \(i\) with higher-income populations. Further, they demonstrate
that \(a_i^{m=\text{public transit}}\) is lower than
\(a_i^{m=\text{car}}\) at many \(i\), arguing that this may put \(i\)
with lower-income populations in a further disadvantage. \(a_i\) and/or
\(a_i^m\) are used to compare relative spatial differences in overall
competitive accessibility and modal competitive accessibility, but
because there is no global maximum, making it is difficult to interpret
the significance between differences in \(a_i^{m}\) values. Questions
such as: what is the impact that competition has on the difference in
\(a_i^m\) values? How does impact vary spatially? And what is the
interpretation of this difference? are left unanswered.

@@

In brief, we define the \emph{spatial availability} at \(i\) ( \(V_{i}\)
) as the proportion of all opportunities in the region \(O\) that are
allocated to location \(i\) from all opportunity destinations \(j\).
\(V_{i}\) is a value of how many opportunities are available to each
location \(i\) out of all the opportunities in the region. The general
formulation of spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}):

@@

Spatial availability measure is introduced in
\citet{soukhovIntroducingSpatialAvailability2023}. The unique feature in
the measure is the balancing factor \(F^t_{ij}\), a proportional
allocation mechanisms, that ensures \(V_i\) calculated for each \(i\)
sums to the total number of opportunities. Through \(F^t_{ij}\), spatial
availability is a \emph{competitive} and \emph{constrained}
accessibility measure that handles the number of opportunities in the
region in a finite way. \(F^t_{ij}\) consists of two components: a
population-based balancing factor \(F^p_{i} = \frac{P_i}{\sum_i P_i}\)
and an impedance-based balancing factor
\(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that, respectively,
allocate opportunities to \(i\) in proportion to the size of the
population at \(i\) (the mass effect) and the cost of reaching
opportunities at \(j\) (the impedance effect).

@@

Consider the following: Figure \ref{fig:Fig1} depicts a region with
population and jobs at three population centers (\(A\), \(B\), \(C\))
and three employment centers (\(1\), \(2\), \(3\)). The population at
each population center is divided into two sub-populations, one using a
faster mode \(z\) and another using a slower mode \(x\), to travel to
employment centers. Population center \(A\) is Suburban: it is closest
to its own relatively large employment center at \(1\), close to the
Urban's equally large employment center \(2\), and has a population that
is smaller than the Urban \(B\) and larger than the Satellite \(C\).
\(B\) has the largest \(x\)-using population, followed by then \(A\),
then \(C\). This synthetic example was inspired by the single-mode
example used in \citet{shenLocationCharacteristicsInnercity1998} and
reconfigured in \citet{soukhovIntroducingSpatialAvailability2023}.

@@

From Figure \ref{fig:Fig2}, it can also be seen that biking and walking
trips are less common than motorized trips at 1\% and 15\% respectively.
The distribution of walking and biking trips appear to be similar to
that of transit trips.

>>>>

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
% \bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\usepackage{forarray}
\usepackage{xstring}
\newcommand{\getIndex}[2]{
  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
}

\setcounter{secnumdepth}{0}

\newcommand{\getAff}[1]{
  \getIndex{#1}{}
}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multimodal Spatial Availability: A Singly-Constrained
Measure of Competitive Accessibility by Multiple
Modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anastasia Soukhov\textsuperscript{\getAff{Department of Earth,
Environment and Society, McMaster University, Canada}},
Javier Tarriño-Ortiz\textsuperscript{\getAff{Universidad Politécnica de
Madrid, Spain}},
Julio A. Soria-Lara\textsuperscript{\getAff{Urban and Regional Planning,
Universidad Politécnica de Madrid, Spain}},
Antonio Páez\textsuperscript{\getAff{Department of Earth, Environment
and Society, McMaster University, Canada}}\\
\bigskip
\bigskip
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
An increasing number of studies within the domain of transportation
planning are concerned with the inequities in accessibility to
opportunities. A dimension of these inequities arises from differences
in access by mode type (e.g., commuting using a car as opposed to
transit). However, methods implemented in current accessibility
literature are lacking within the context of multimodal analysis. This
paper presents an extension of spatial availability, a
singly-constrained competitive accessibility measure, for the context of
multimodal accessibility analysis. We first illustrate the features of
spatial availability that lend itself to multimodal analysis. We then
demonstrate its use on the case study of Low Emission Zones in Madrid
(Spain) and highlight how this policy intervention changes the
accessibility of populations using different modes. In summary, spatial
availability can be used to create and interpret multimodal policy
intervention scenarios unlike previous methods: this creation and
interpretation can help regions envision a more sustainable and
equitable access-to-opportunity landscape.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

An increasing number of studies within the domain of transport are
concerned with the inequities in accessibility to opportunities. A
dimension of these inequities arise from differences in access by mode
type (e.g., the number of work opportunities that can be reached using a
car as opposed to transit in a city). However, methods assessing
multimodal accessibility in the literature fall short as aspects of
competition for opportunities and the explicit methodological
acknowledgement of opportunities being \emph{finite} are lacking. In
this vein, this paper presents an extension of \emph{spatial
availability}, a singly-constrained competitive accessibility measure,
for the context of multimodal accessibility analysis. We first
illustrate the features of spatial availability that lends itself to
multimodal analysis. We then demonstrate its use on the case study of
Low Emission Zones in Madrid (Spain) and highlight how this policy
intervention changes the accessibility for locations where populations
use different modes. In summary, spatial availability can be used to
create and interpret multimodal policy intervention scenarios unlike
previous methods: this creation and interpretation can help regions
envision a more sustainable and equitable access-to-opportunity
landscape by better identifying differences in accessibility afforded by
different modes.

@@

Implementing urban policies that re-shape cities through accessibility
gains (i.e., the \emph{potential to interact} with opportunities as a
result of land-use mix and transport systems as originally defined by
{[}1{]}) have been widely applied within the transportation literature
and is increasingly discussed by planners {[}2--5{]}. An important
challenge in the identification of interventions that equitably
transform cities is the effective evaluation of \emph{trade-offs}:
cities are complex and dynamic ecologies, and advantaging one component
of the city can disadvantage another area, population, or sub-component
In this way, policy evaluation should take a \emph{systems} approach
{[}6{]}. One way of considering systems is from the perspective of the
\emph{finite}. As an illustration, consider the amount of transport
space within a city: the amount is typically finite so re-allocating
road space away from one mode directly impacts the performance of the
others (see the literature on road space reallocation e.g., {[}7{]}).
Evaluating policy impacts in the context of \emph{finity} provides a way
to contextualize the balance of trade-offs that the citizens of a city
should tolerate.

From the perspective of urban transport systems, location-based
accessibility measures have been used in the context of policy
evaluation. For instance, {[}8{]} assesses the transit accessibility
gains to healthcare and employment opportunities for disadvantaged
neighbourhood in Columbus, Ohio, USA after the transit system's
re-design and introduction of a rapid bus system. However, a limitation
of this study, like others that implement accessibility measures, is
they do not calculate results under a \emph{constrained} framework i.e.,
one of \emph{finity}. The citizens of Columbus should experience
quantitative accessibility gains - but is it at the expense of access
using other modes? As another example, {[}9{]} implements a modified
cumulative opportunity measure to assess differences between private
vehicle and transit system accessibility to jobs in Melbourne, but a
similar question remains: does the accessibility afforded to the private
vehicle using population come at the expense of accessibility losses to
transit users?

{[}8{]} and {[}9{]} both use \emph{non-competitive} accessibility
measures. There is a branch of location-based accessibility measures
that do incorporate the effect of competition for opportunities by the
population in the region. However, we argue that these existing methods
fall short in acknowledging the \emph{finity} of opportunities. For
instance, {[}10{]} applies a competitive measure, two-step floating
catchment approach (2SFCA), for the case of access to healthcare
services in Florida for both a multimodal network and a single modal
network. While the differences in modal access are discussed, the
question of how the advantage in access afforded by one mode over
another impacts access for different mode users is unanswered.

This question of how much one mode-using population can access at the
expense of another mode-using population is a pertinent equity question
in the evaluation of policy scenarios that are multimodal. For instance,
consider the impact of a low emission zone (LEZ). LEZ is a policy of
spatial and modal discrimination: the circulation of vehicles that are
excessively polluting are restricted in specific regions. In the
recognition that opportunities are finite, the implementation of a LEZ
explicitly reduces the access that the population using polluting
vehicles has to opportunities. This restriction allows the population
using other more sustainable modes to potentially have a higher level of
access than before the LEZ implementation. This evaluation is especially
urgent as LEZ are currently in effect in cities globally; their
reception has been mixed {[}11{]} and may be having negative impacts on
disadvantaged populations who have become mobility-restricted
{[}12,13{]}. Measures that evaluate the accessibility of modes given
both \emph{constrained} and \emph{competitive} considerations are
lacking in the literature, but are needed, to evaluate such policy
interventions impact on accessibility.

In {[}14{]}, we introduce spatial availability, a type of location-based
accessibility measure that is both \emph{constrained} and
\emph{competitive}. In this paper, we extend the spatial availability
measure into a multimodal framework and explore its use in answering the
question outlined: \emph{``given opportunities are finite, how many are
available to a given location depending on the mode used?''.} The answer
to this question quantifies how many opportunities can be accessed,
considering competition, for different modes. To foreground this
exploration, in Section 2, we discuss short falls of a few existing
location-based measures in comparison to spatial availability through a
synthetic example. In Section 3, the spatial availability of an
empirical example of the LEZ in the city of Madrid, Spain is calculated.
We demonstrate how the restriction of car circulation could have
impacted the spatial availability of opportunities for each
sub-population using transit, cycling and walking modes. In Section 4,
we provide concluding remarks on the strengths of the use of spatial
availability as a multimodal accessibility measure and potential future
uses in policy planning scenarios.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are a product of the relationship between land-use
and transport systems. Arguably the most commonly used location-based
measured are cumulative opportunity measures and weighted cumulative
opportunity measures {[}2{]}. These measures weight the opportunities
that can be potentially interacted with from origin \(i\) to destination
\(j\) based on some sort of travel cost function (e.g., travel time,
fare, travel distance) otherwise known as a travel impedance function
\(f^{m}(c^m_{ij})\). Many weighted cumulative opportunities (often
referred to as the gravity-based measure) originate from the measure
proposed by {[}1{]}, which can take the following multimodal form:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
have mode-specific travel costs \((c_{ij}^m)\) and travel impedance
functions \(f^m(\cdot)\).

The Hansen-type measure does not consider competition between modes nor
is it constrained. As an example, the work of {[}15{]} uses the
Hansen-type measure to measure the potential interaction with retail
locations using walking, public transit, and car modes \(m\). \(S_i^m\)
is the sum of retail locations \(j\) that can potentially be interacted
with under the travel impedance as calculated for each \(i\) and \(m\).
In other words, each \(i\) has three \(S_i\) values, one per \(m\). In
this work, they demonstrate that the car mode has the highest
\(S_i^{m=car}\) values in the majority of \(i\), i.e., populations using
a car can potentially interact with the most retail opportunities than
populations using other modes. However, the higher \(S_i^{m=car}\)
values are not a result of lower \(S_i^{m}\) values for other modes: it
is not assumed that car-using populations potentially accessing more
opportunities take away potential opportunities for other populations
within the measure (no consideration for competition). This measure is
also not constrained: there is no global maximum for \(S_i\) or
\(S_i^m\) values, they are presented as a population normalized
accessibility index. This makes the interpretation of the `potentially
interacted opportunities' relative to the region, making comparisons of
the results across different regions challenging.

However, opportunities in a region can be considered finite. There are
only so many school-seats, hospital capacity, jobs, etc., in a region
and if one person interacts with an opportunity at a given time, it is
taken. As such, if one person is advantaged and has the ability to reach
more opportunities through a lower travel-cost mode, than they have more
opportunities to potentially interact with more opportunities than other
people. From the other perspective, their are fewer opportunities left
to be potentially interacted with for populations using higher
travel-cost modes. In this way, populations using modes with a higher
travel impedance are at a higher access disadvantage than populations
using lower travel impedance modes. This recognition is the motivation
behind integrating \emph{competition} for opportunities within
multimodal accessibility measures. Arguably one of the most popular
competitive location-based accessibility measures is the two-step
floating catchment area (2SFCA) approach popularized by {[}16{]} who
simplified the approach proposed by {[}17{]} (with similar
considerations for competition in {[}18{]} and {[}19{]}).

@@

To illustrate, {[}20{]} calculates \(a_i^m\) to jobs for different
income-group populations in Shenzhen (China) using
\(m = \text{public transit}\) and \(m={car}\). They demonstrate that
\(i\) with low-income populations have lower \(a_i^m\) than \(i\) with
higher-income populations. Further, they demonstrate that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many \(i\), arguing that this may put \(i\) with lower-income
populations in a further disadvantage. \(a_i\) and/or \(a_i^m\) are used
to compare relative spatial differences in overall competitive
accessibility and modal competitive accessibility, but because there is
no global maximum, making it is difficult to interpret the significance
between differences in \(a_i^{m}\) values. Questions such as: what is
the impact that competition has on the difference in \(a_i^m\) values?
How does impact vary spatially? And what is the interpretation of this
difference? are left unanswered.

@@

In brief, we define the \emph{spatial availability} at \(i\) (\(V_{i}\))
as the proportion of all opportunities in the region \(O\) that are
allocated to location \(i\) from all opportunity destinations \(j\).
\(V_{i}\) is a value of how many opportunities are available to each
location \(i\) out of all the opportunities in the region. The general
formulation of spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}):

@@

The spatial availability measure is introduced in {[}14{]}. Spatial
availability's unique feature is the balancing factor \(F^t_{ij}\), a
proportional allocation mechanisms, that ensures the \(V_i\) calculated
for each \(i\) sums, across all \(i\) in the region, to equal the total
number of opportunities in the region. As such, spatial availability is
a \emph{competitive} and \emph{constrained} accessibility measure as
\(F^t_{ij}\) handles the number of opportunities in the region in a
finite way (proportional allocation). \(F^t_{ij}\) consists of two
components: a population-based balancing factor
\(F^p_{i} = \frac{P_i}{\sum_i P_i}\) and an impedance-based balancing
factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
respectively, allocate opportunities to \(i\) in proportion to the size
of the population at \(i\) (the mass effect) and the cost of reaching
opportunities at \(j\) (the impedance effect).

@@

Consider the following: Figure \ref{fig:Fig1} depicts a region with
population and jobs at three population centers (\(A\), \(B\), \(C\))
and three employment centers (\(1\), \(2\), \(3\)). The population at
each population center is divided into two sub-populations, one using a
faster mode \(z\) and another using a slower mode \(x\), to travel to
employment centers. Population center \(A\) is Suburban: it is closest
to its own relatively large employment center at \(1\), close to the
Urban's equally large employment center \(2\), and has a population that
is smaller than the Urban \(B\) and larger than the Satellite \(C\).
\(B\) has the largest \(x\)-using population, followed by then \(A\),
then \(C\). This synthetic example was inspired by the single-mode
example used in {[}17{]} and reconfigured in {[}14{]}.

@@

From Figure \ref{fig:Fig3}, it can also be seen that biking and walking
trips are less common than motorized trips at 1\% and 15\% respectively.
Noteably, there is a positive trend between the populations of walking
and biking trips in zones and populations of transit trips. This
positive trend is higher than for car trip populations.
